Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-07 content.tex
Example is more readable now.

\begin{center}
  $\begin{array}{l}\includegraphics[scale=0.4195]{example1}\end{array}$
  $\Rightarrow$
  $\begin{array}{l}\includegraphics[scale=0.4195]{example2}\end{array}$
  $\Rightarrow$
  $\begin{array}{l}\includegraphics[scale=0.4195]{example3}\end{array}$
\end{center}

{\small
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\# & $\mstore$ & $\mctx$ & $\mkont$ & $\mktab$ & $\mmemo$ \\
\hline
0 & $[{\tt id} \mapsto \set{(\slam{\mvar}{\mvar},\bot)}]$ & $(({\tt x}, \iota), \mstore_1)$ & $\mkont_0$ & $[\mctx_0 \mapsto \set{\mkont_0}]$ & $[\mctx_0 \mapsto \set{0}]$ \\
1 & $\mstore_0[{\tt \mvar} \mapsto \set{0}]$ & $(({\tt x}, \iota), \mstore_3)$ & $\krt{\mctx_0}$ & $\mktab_0[\mctx_1 \mapsto \set{\mkont_2}]$ & $\mmemo_0[\mctx_1 \mapsto \set{0,1}]$ \\
2 & $\mstore_1[{\tt y} \mapsto \set{0}]$ & - & $\mathbf{lt}({\tt z}, {\tt (}\le\ {\tt y}\ {\tt z)})$ & - & - \\
3 & $\mstore_2[{\tt x} \mapsto \set{0,1}]$ & - & $\krt{\mctx_1}$ & - & - \\
4 & $\mstore_3[{\tt z} \mapsto \set{0,1}]$ & - & $\kmt$ & - & - \\
\hline
\end{tabular}}
Where $\mkont_0 = \mathbf{lt}({\tt y}, {\tt (let\ ([z\ (id\ 1)])\ (}\le\ {\tt y}\ {\tt z))}, \iota)$

% %% FIXME stores in memo-table
% \begin{enumerate}
% \item{\texttt{(id 0)} steps to \texttt{x} at $\krt{((\texttt{x}, [\texttt{x}\mapsto\texttt{x}]), \mstore_0)}$ with $\mstore_1 =
%     \mstore_0[\texttt{x} \mapsto \set{0}]$ and
%     (let $\mctx_1 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_1)$,
%          $\mkont_1 = \klet{\texttt{y}, \texttt{(let ([z (id 1)]) ($\le$ y z))}, [\texttt{id}\mapsto{\texttt{id}}]}$)
%     $\mktab_1 = [\mctx_0 \mapsto \set{\mkont_0}]$}
% \item{\texttt{0} at $\krt{\mctx_1}$ steps to $0$ at $\mkont_1$ and $\mmemo_1 = [\mctx_1 \mapsto \set{0}]$}
% \item{\texttt{0} at $\mkont_1$ steps to \texttt{(let ([z (id 1)]) ($\le$ y z))} and $\mstore_2 = \mstore_1[\texttt{y} \mapsto \set{0}]$}
% \item{\texttt{(let ([z (id 1)]) ($\le$ y z))} steps to \texttt{(id 1)} at \\
%       $\klet{\texttt{z}, \texttt{($\le$ y z)}, [\texttt{id} \mapsto \texttt{id}, \texttt{y} \mapsto \texttt{y}]}$
%       (call this $\mkont_2$).}
% \item{\texttt{(id 1)} steps to $x$ with $\mstore_3 = \mstore_2[\texttt{x} \mapsto \set{0,1}]$ and
%       (let $\mctx_3 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_3)$) $\mktab_3 =\mktab_1[\mctx_3 \mapsto \set{\mkont_2}]$.}
% \item{\texttt{0} or \texttt{1} at $\krt{\mctx_3}$ steps to \texttt{0} or \texttt{1} at $\mkont_2$ and $\mmemo_3 = \mmemo_1[\mctx_3 \mapsto \set{0,1}]$.}
% \item{\texttt{z} gets bound to $\set{0,1}$, and \texttt{($\le$ y z)} evaluates to true.}
% \end{enumerate}

@@

The astute reader may find this definition is dubious, since there is a possibility for unguarded corecursion here.
%
Interpreting a function with ill-founded recursion can lead to a table such that $\krt{\mctx} \in \mktab(\mctx)$.
%
The abstractions that we impose on our data might also cause this situation, even though the concrete execution might always terminate.
%
Our interpretation of $\stepto$ is in a fixed-point computation, so circularities are not a problem for termination, but we augment the definition here to guard against circularities in our proofs:

\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt, G}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont', \emptyset}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont', G}} \\
  \inferrule{\mctx \notin G  \\
             \mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ, \set{\mctx}\cup G}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}, G}}  
\end{mathpar}
Consider later uses of $\mathit{unroll}$ to implicitly give $\emptyset$ for $G$.

@@

The different tables we have encode information about execution history that we prove is invariant.
\begin{align*}
  \inv(\tpl{\mpoint, \mstore, \mkont, \mktab, \mmemo}) &=
          \dom(\mmemo) \subseteq \dom(\mktab) \\
  &\wedge \forall (\mpoint',\mstore') \in \dom(\mktab),
                  \mkont^\circ \in \mktab(\mpoint',\mstore'),
                  \mkont' \in \unroll{\mktab}{\mkont^\circ}. \\   
  &\qquad \phi_{\reachable} \wedge \phi_{\memo} \\
  \text{where } \phi_{\reachable} &=
  \exists \mkont^* \in \unroll{\mktab}{\mkont}.
    \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mpoint, \mstore, \mkont^*} \\
  \phi_{\memo} &=
 \forall (\mval, \mstore'') \in \mmemo(\mpoint',\mstore'). \\
  &\qquad \exists \mtrace\equiv\tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mval,\mstore'', \mkont'}.
     \hastail(\mtrace,\mkont') \\
\end{align*}

\begin{theorem}[Equivalence with global tables]\label{thm:global}
  Tabular semantics has a bisimulation with memo/k-table-widened semantics.
\end{theorem}

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers (\autoref{thm:gc-pointers}), equal length traces are equal up to GC (in the concrete) (\autoref{thm:gc-concrete}), and abstractions are sound with respect to collected states (\autoref{thm:gc-sound}).
%
\begin{theorem}\label{thm:gc-pointers}
  All addresses in $\Gamma(\mstate)$ are in the domain of $\Gamma(\mstate).\mstore$.
\end{theorem}

@@

%%
The second of these is to ameliorate a problem they call ``fake rebinding.''
%
That is, since bindings in the abstract represent several values, we don't want to reference a variable \texttt{x} in two different places and have it resolve to two different values.
%
In AAM, a variable reference non-deterministically steps to all possible values associated with that variable.
%
Here we want to say that once \texttt{x} is considered to stand for value $v$, then all subsequent references of \texttt{x} should be $v$.
%
If they aren't, it looks as if \texttt{x} was rebound; it hasn't, and thus it is a ``fake rebinding.''
%
CFA2 does not step to all values on variable reference, but instead carries all its values around in superposition until they need to be observed at, say, a function call.
%
We give a simplified semantics that is more along the AAM style, but CFA2's approach can easily be recovered from it\footnote{Our Redex model implements the fake rebinding strategy CFA2 itself employs}.
%%

@@

As the programming world continues to embrace behavioral values, it becomes more important to import the powerful techniques pioneered by the first-order analysis and model checking communities.
%
CFA2 and PDCFA paved the way, and in large part inspired this work.
%
It is our view that systematic approaches to applying the techniques are pivotal to scaling them to ``real languages.''
%
We believe that the recipe that this paper set forth is a step in that direction.
%
That is, make continuation tables keyed with enough context, and memoize at the introduced indirection points.
%
The result in a language with well-bracketed control is a ``pushdown analysis'' using summarization.
%
In a language without well-bracketed control, we are not chained to a pushdown automaton as the target of the approximation, so the techniques still apply and give better precision than regular methods.
%%

%%
Our goal in the future is to show that this technique is even more widely applicable than shift and reset.
%
We conjecture that the same recipe will apply to the most intricate control operators in production languages such as in \citet{ianjohnson:Flatt:2007:ADC:1291151.1291178}.
%
The control structures there are difficult to model even with AAM's techniques due to the ability to capture and compose continuations with arbitrarily many prompts, but once we can tackle AAM, there should be a straightforward route to extending it with the pushdown techniques of this paper.
%
We have an unproven prototype of the stack inspection capabilities provided by the {\tt with-continuation-mark}, {\tt current-continuation-marks} and {\tt continuation-mark-set->list} primitives, where a ``context'' then also includes what is the (approximated) result of {\tt current-continuation-marks}, since functions can inspect only this amount of the stack.
%%

>>>>

\begin{center}
  \begin{minipage}{0.55\linewidth}
    \begin{center}
      $\begin{array}{l}\includegraphics[scale=0.65]{example1}\end{array}$ \\
      $\Downarrow$ \\
      $\begin{array}{l}\includegraphics[scale=0.65]{example2}\end{array}$ \\
      $\Downarrow$ \\
      $\begin{array}{l}\includegraphics[scale=0.65]{example3}\end{array}$
    \end{center}
  \end{minipage}
  \begin{minipage}{0.43\linewidth}
    \begin{center} {\small
        \begin{tabular}{|l|c|c|}
          \hline
          \# & $\mstore$ & $\mkont$ \\
          \hline
          0 & $[{\tt id} \mapsto \set{(\slam{\mvar}{\mvar},\bot)}]$ & $\mkont_0$ \\
          1 & $\mstore_0[{\tt \mvar} \mapsto \set{0}]$ & $\krt{\mctx_0}$ \\
          2 & $\mstore_1[{\tt y} \mapsto \set{0}]$ & $\mathbf{lt}({\tt z}, {\tt (}\le\ {\tt y}\ {\tt z)})$ \\
          3 & $\mstore_2[{\tt x} \mapsto \set{0,1}]$ & $\krt{\mctx_1}$ \\
          4 & $\mstore_3[{\tt z} \mapsto \set{0,1}]$ & $\kmt$ \\
          \hline
        \end{tabular}} \\[10pt]
      {\small
        \begin{tabular}{|l|c|}
          \hline
          \# & $\mctx$ \\
          \hline
          0  & $(({\tt x}, \iota), \mstore_1)$ \\
          1  & $(({\tt x}, \iota), \mstore_3)$ \\
          \hline
        \end{tabular}}
      \\[10pt]
      {\small
        \begin{tabular}{|l|c|c|c|}
          \hline
          \# & $\mktab$ & $\mmemo$ \\
          \hline
          0  & $[\mctx_0 \mapsto \set{\mkont_0}]$ & $[\mctx_0 \mapsto \set{0}]$ \\
          1  & $\mktab_0[\mctx_1 \mapsto \set{\mkont_2}]$ & $\mmemo_0[\mctx_1 \mapsto \set{0,1}]$ \\
          \hline
        \end{tabular}}    
    \end{center}
  \end{minipage}
\end{center}
Where $\mkont_0 = \mathbf{lt}({\tt y}, {\tt (let\ ([z\ (id\ 1)])\ (}\le\ {\tt y}\ {\tt z))}, \iota)$
% %% FIXME stores in memo-table
% \begin{enumerate}
% \item{\texttt{(id 0)} steps to \texttt{x} at $\krt{((\texttt{x}, [\texttt{x}\mapsto\texttt{x}]), \mstore_0)}$ with $\mstore_1 =
%     \mstore_0[\texttt{x} \mapsto \set{0}]$ and
%     (let $\mctx_1 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_1)$,
%          $\mkont_1 = \klet{\texttt{y}, \texttt{(let ([z (id 1)]) ($\le$ y z))}, [\texttt{id}\mapsto{\texttt{id}}]}$)
%     $\mktab_1 = [\mctx_0 \mapsto \set{\mkont_0}]$}
% \item{\texttt{0} at $\krt{\mctx_1}$ steps to $0$ at $\mkont_1$ and $\mmemo_1 = [\mctx_1 \mapsto \set{0}]$}
% \item{\texttt{0} at $\mkont_1$ steps to \texttt{(let ([z (id 1)]) ($\le$ y z))} and $\mstore_2 = \mstore_1[\texttt{y} \mapsto \set{0}]$}
% \item{\texttt{(let ([z (id 1)]) ($\le$ y z))} steps to \texttt{(id 1)} at \\
%       $\klet{\texttt{z}, \texttt{($\le$ y z)}, [\texttt{id} \mapsto \texttt{id}, \texttt{y} \mapsto \texttt{y}]}$
%       (call this $\mkont_2$).}
% \item{\texttt{(id 1)} steps to $x$ with $\mstore_3 = \mstore_2[\texttt{x} \mapsto \set{0,1}]$ and
%       (let $\mctx_3 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_3)$) $\mktab_3 =\mktab_1[\mctx_3 \mapsto \set{\mkont_2}]$.}
% \item{\texttt{0} or \texttt{1} at $\krt{\mctx_3}$ steps to \texttt{0} or \texttt{1} at $\mkont_2$ and $\mmemo_3 = \mmemo_1[\mctx_3 \mapsto \set{0,1}]$.}
% \item{\texttt{z} gets bound to $\set{0,1}$, and \texttt{($\le$ y z)} evaluates to true.}
% \end{enumerate}

@@

The astute reader may find this definition dubious, since there is a possibility for unguarded corecursion here.
%
Interpreting a function with ill-founded recursion can lead to a table such that $\krt{\mctx} \in \mktab(\mctx)$.
%
The abstractions that we impose on our data might also cause this situation, even though the concrete execution might always terminate.
%
Our interpretation of $\stepto$ is in a fixed-point computation, so circularities are not a problem for termination, but we augment the definition here to guard against circularities in our proofs:

\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt, G}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont', \emptyset}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont', G}} \\
  \inferrule{\mctx \notin G  \\
             \mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ, \set{\mctx}\cup G}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}, G}}  
\end{mathpar}

%%
The different tables we have encode information about execution history that we prove is invariant.
\begin{align*}
  \inv(\tpl{\mpoint, \mstore, \mkont, \mktab, \mmemo}) &=
          \dom(\mmemo) \subseteq \dom(\mktab) \\
  &\wedge \forall (\mpoint',\mstore') \in \dom(\mktab),
                  \mkont^\circ \in \mktab(\mpoint',\mstore'),
                  \mkont' \in \unroll{\mktab}{\mkont^\circ, \emptyset}. \\   
  &\qquad \phi_{\reachable} \wedge \phi_{\memo} \\
  \text{where } \phi_{\reachable} &=
  \exists \mkont^* \in \unroll{\mktab}{\mkont,\emptyset}.
    \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mpoint, \mstore, \mkont^*} \\
  \phi_{\memo} &=
 \forall (\mval, \mstore'') \in \mmemo(\mpoint',\mstore'). \\
  &\qquad \exists \mtrace\equiv\tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mval,\mstore'', \mkont'}.
     \hastail(\mtrace,\mkont') \\
\end{align*}

In order to prove this, we need a lemma that allows us to plug in any continuation for memoized results.

@@

Finally, we can show that even if the tables are shared amongst all states, they are precise enough to not interfere with the results.
\begin{theorem}[Equivalence with global tables]\label{thm:global}
  Tabular semantics has a bisimulation with memo/k-table-widened semantics.
\end{theorem}

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers (\autoref{thm:gc-pointers}), equal length traces are equal up to GC (in the concrete) (\autoref{thm:gc-concrete}), and abstractions are sound with respect to collected states (\autoref{thm:gc-sound}).
%
Technically since this semantics inspects the touched addresses of the continuation, the set of touched addresses should be considered part of the ``context'' in the $\mathbf{rt}$ continuations.
%
This extra context can be overlooked, since that means the computed touched addresses will be an overapproximation of what is already sound.
%
More touched addresses means a less effective GC, but this is an acceptable tradeoff for a possibly smaller state space.
%
\begin{theorem}\label{thm:gc-pointers}
  All addresses in $\Gamma(\mstate)$ are in the domain of $\Gamma(\mstate).\mstore$.
\end{theorem}

@@

%%
The second of these is to ameliorate a problem they call ``fake rebinding.''
%
That is, since bindings in the abstract represent several values, we don't want to reference a variable \texttt{x} in two different places and have it resolve to two different values.
%
In AAM, a variable reference non-deterministically steps to all possible values associated with that variable.
%
Here we want to say that once \texttt{x} is considered to stand for value $v$, then all subsequent references of \texttt{x} should be $v$.
%
If they are not all considered $v$, it looks as if \texttt{x} was rebound; it hasn't, and thus it is a ``fake rebinding.''
%
CFA2 does not step to all values on variable reference, but instead carries all its values around in superposition until they need to be observed at, say, a function call.
%
We give a simplified semantics that is more along the AAM style, but CFA2's approach can easily be recovered from it\footnote{Our Redex model implements the fake rebinding strategy CFA2 itself employs}.
%%

@@

As the programming world continues to embrace behavioral values, it becomes more important to import the powerful techniques pioneered by the first-order analysis and model checking communities.
%
CFA2 and PDCFA paved the way, and in large part inspired this work.
%
It is our view that systematic approaches to applying the techniques are pivotal to scaling them to production programming languages.
%
We believe that the recipe that this paper set forth is a step in that direction.
%
That is, make continuation tables keyed with enough context, and memoize at the introduced indirection points.
%
The result in a language with well-bracketed control is a ``pushdown analysis'' using summarization.
%
In a language without well-bracketed control, we are not chained to a pushdown automaton as the target of the approximation, so the techniques still apply and give better precision than regular methods.
%%

%%
Our goal in the future is to show that this technique is even more widely applicable than shift and reset.
%
We conjecture that the same recipe will apply to the most intricate control operators in production languages such as in \citet{ianjohnson:Flatt:2007:ADC:1291151.1291178}.
%
The control structures there are difficult to model even with AAM's techniques due to the ability to capture and compose continuations with arbitrarily many prompts, but once we can tackle AAM, there should be a straightforward route to extending it with the pushdown techniques of this paper.
%
We have an unproven prototype of the stack inspection capabilities provided by the {\tt with-continuation-mark}, {\tt current-continuation-marks} and {\tt continuation-mark-set->list} primitives, where a ``context'' then also includes what is the result of {\tt current-continuation-marks}, since functions can inspect only this amount of the stack.
%%
